distr shiftate + troncate

### 2016-09-15 docs/article-shift+trunc.Rmd
articolo distr shiftate e troncate

R programming functional nature allows us programmer to implement and manipulate functions as we need. Functional programming is a style of programming that emphasizes the evaluation of expressions rather than the execution of commands. It has a great power and lots of properties that combined together can allows us to produce results with the most efficient computational method. In this article we will consider the functional factory method proposed by Andrea Spanò in "R package qdist: A functional programming approach to truncated probability ditributions", and that I used in "R package qshift: A functional programming approach to shifted probability distributions", that produces functions for truncated and shifted distributions, respectively. [`qdist`](https://github.com/andreaspano/qdist) and [`qshift`](https://github.com/veronicagiro/qshift) packages are available on Github:

@@

We will demonstrate that it is possible to produce functions for truncated and shifted or for shifted and truncated distributions only combining the functions of the two packages.

@@

<!--
pensare ad un modo diverso per spiegare la cosa perchè da questi grafici nn si capisce niente
-->
## Other Examples

Let us see how it the other distribution functions works with other distributions.

Let us consider Poisson distribution and distribution function.

First of all functions for truncated distribution and for shifted distributio have to be defined:

```{r}
dtpois <- dtruncate("pois")
ptpois <- ptruncate("pois")
qtpois <- qtruncate("pois")
rtpois <- rtruncate("pois")

dspois <- dshift("pois")
pspois <- pshift("pois")
qspois <- qshift("pois")
rspois <- rshift("pois")
```

>>>>

R programming functional nature allows us programmer to implement and manipulate functions as we need. Functional programming is a style of programming that emphasizes the evaluation of expressions rather than the execution of commands, so these functions produce results that depend only on their inputs and not on the program state. This style has a great power and includes lots of tools that combined together can allows us to produce results with the most efficient computational method. One of them is the functional factory, which is a factory for making new functions. It allows us to use closures, which are functions together with a referencing environment, to write specific functions that, in turn, can be used to generate new functions. So we can have a double layer of development: a first layer that is used to do all the complex work in common to all functions and a second layer that defines the details of each function. This fundamentals was used to develop the computational method proposed by Andrea Spanò in "R package qdist: A functional programming approach to truncated probability ditributions", and that I used in "R package qshift: A functional programming approach to shifted probability distributions", that produces functions for truncated and shifted distributions, respectively.  [`qdist`](https://github.com/andreaspano/qdist) and [`qshift`](https://github.com/veronicagiro/qshift) packages are available on Github:

@@

In the previous citated articles, we dealt separately shifted and truncated distributions. But in the real life, we sometimes work with random variables arising from more than one transformation. So in this article, we will demonstrate that it is possible to produce functions for truncated and shifted or for shifted and truncated distributions only combining the functions of the two packages thanks to functional programming approach used to built them.

<!--

@@

pensare ad un modo diverso per spiegare la cosa perchè da questi grafici nn si capisce niente
-->

R provides probability, density, quantile and random number generation functions in `stats` package. The functions, that we refer to, share a stable naming convention both for the names of the functions and in the first argument of these functions. In particular:

* `p<dist>(q)` with `q` vector of quantiles, identifies probability distribution functions
* `d<dist>(x)` with `x` vector of quantiles, identifies density functions
* `q<dist>(p)` with `p` vector of probabilities, identifies quantile functions
* `r<dist>(n)` with `n` number of observations, identifies random number generation

where `<dist>` indicates the distribution family.

As a result we have: `pnorm(q)`, `dnorm(x)`, `qnorm(p)` and `rnorm(n)` for normal distribution and similarly for all other distributions.

Andrea Spanò proposed in `qdist` a functional programming method that computes truncated distributions functions for each distribution chosen without changing the well structured programming style and syntax of previous functions.

The computational methodology proposed in `qdist` is based on two R key concepts:

As `dnorm()` is the function that computes the density function for a normal distribution, conventionally named `norm`, `dtnorm()` is the function that computes the density function for a truncated normal distributions, named `tnorm`.

Suppose we want to generate the density functions for a **shifted truncated** normal distribution. We have to entrust to `qshift` package. `qshift` package shares the same logic and the same methodology of `qdist`. The difference is that the functions are designed to compute functions for shifted distributions.
In particular, `qshift` package includes the following four functions:

* `dshift()`, for density function
* `pshift()`, for distribution function
* `qshift()`, for quantile function
* `rshift()`, for random generator function

```{r, message=FALSE}
require(qshift)
dstnorm <- dshift("tnorm")
```

Let us graphically compare the results of `dnorm()`, `dtnorm()` and `dstnorm()` functions:

```{r}
x <- seq(6, 14, len = 100)
std_density <- dnorm(x = x, mean = 10, sd = 1)
trunc_density <- dtnorm(x = x, mean = 10, sd = 1, L = 9, U = 11)
shift_trunc_density <- dstnorm(x = x, mean = 10, sd = 1, L = 9, U = 11, shift = 1)
```

```{r}
df <- data.frame(x=x, 
                 type = factor(c(rep("std_density", times=length(x)),
                    rep("trunc_density", times=length(x)),
                    rep("shift_trunc_density", times=length(x))),
                    levels = c("std_density", "trunc_density", "shift_trunc_density")), 
                  value = c(std_density, trunc_density, shift_trunc_density))
require(ggplot2)
pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("std_density" = "cyan3", 
                                "trunc_density" = "#89FF99", 
                                "shift_trunc_density" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle(paste("Density plot for non-transformed, truncated", "and shifted truncated Normal distributions", sep = "\n")) +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

The inverse procedure generates the density function for a **truncated shifted** normal distribution.

Starting from `dshift()` function of `qshift` package, we define `dsnorm()`, the density function of `snorm`, which is the shifted normal distribution:
```{r}
dsnorm <- dshift("norm")
```

Let us graphically represent the output of the three functions:

```{r}
std_density <- dnorm(x = x, mean = 10, sd = 1)
shift_density <- dsnorm(x = x, mean = 10, sd = 1, shift = 1)
trunc_shift_density <- dtsnorm(x = x, mean = 10, sd = 1, L = 9, U = 11, shift = 1)
```

```{r}
df <- data.frame(x=x, 
                 type = factor(c(rep("std_density", times=length(x)),
                    rep("shift_density", times=length(x)),
                    rep("trunc_shift_density", times=length(x))),
                    levels = c("std_density", "shift_density", "trunc_shift_density")), 
                  value = c(std_density, shift_density, trunc_shift_density))

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("std_density" = "cyan3", 
                                "shift_density" = "#89FF99", 
                                "trunc_shift_density" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle(paste("Density plot for non-transformed, shifted and","truncated shifted Normal distributions", sep = "\n")) +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

```{r}
df <- data.frame(x=x, 
                 type = factor(c(rep("std_density", times=length(x)),
                    rep("trunc_shift_density", times=length(x)),
                    rep("shift_trunc_density", times=length(x))),
                    levels = c("std_density", "trunc_shift_density", "shift_trunc_density")), 
                  value = c(std_density, trunc_shift_density, shift_trunc_density))

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("std_density" = "cyan3", 
                                "trunc_shift_density" = "#89FF99", 
                                "shift_trunc_density" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle(paste("Density plot for non-transformed, truncated-shifted", "and shifted-truncated Normal distribution", sep="\n")) +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

Looking at the figure we can see that a shifted truncated distribution is different from a truncated shifted distribution.

pl <- ggplot(data = df, mapping = aes(x=value, group=type, fill=type)) +
        geom_density(alpha=0.6) +
        scale_fill_manual(values = c("std_rand" = "cyan3",
                              "st_rand" = "#89FF99",
                              "ts_rand" = "#FF89B4")) +
        ggtitle("Density plot for shifted and not-shifted Normal distributions") +
        theme(legend.title=element_blank(), legend.text= element_text(size=10),
        plot.title = element_text(size=14), axis.title=element_text(size=12))

## Other examples

Let us consider Poisson distribution and define the distribution function for shifted and truncated Poisson distribution.

```{r}
dspois <- dshift("pois")
pspois <- pshift("pois")
qspois <- qshift("pois")
rspois <- rshift("pois")

dtpois <- dtruncate("pois")
ptpois <- ptruncate("pois")
qtpois <- qtruncate("pois")
rtpois <- rtruncate("pois")
```

* Density function of shifted truncated and truncated shifted Poisson distribution.

```{r}
# truncated shifted
dtspois <- dtruncate("spois")
# shifted truncated 
dstpois <- dshift("tpois")
```

```{r}
x <- 1:10
std_density <- dpois(x = x, lambda = 3)
shift_trunc_density <- dstpois(x = x, lambda = 3, shift = 1, L = 2, U = 7)
trunc_shift_density <- dtspois(x = x, lambda = 3, L = 2, U = 7, shift = 1)
```

```{r}
df <- data.frame(x=x, 
                 type = factor(c(rep("std_density", times=length(x)),
                    rep("shift_trunc_density", times=length(x)),
                    rep("trunc_shift_density", times=length(x))),
                    levels = c("std_density", "shift_trunc_density", "trunc_shift_density")), 
                  value = c(std_density, shift_trunc_density, trunc_shift_density))

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("std_density" = "cyan3", 
                                "shift_trunc_density" = "#89FF99", 
                                "trunc_shift_density" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle(paste("Density plot for non-transformed, shifted and","truncated shifted Normal distributions", sep = "\n")) +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

* Probability function of shifted truncated and truncated shifted Poisson distribution.

```{r}
# truncated shifted
ptspois <- ptruncate("spois")
# shifted truncated 
pstpois <- pshift("tpois")
```

```{r}
std_distr <- ppois(q = x, lambda = 3)
st_distr <- pstpois(q = x, lambda = 3, L = 2, U = 7, shift = 1)
ts_distr <- ptspois(q = x, lambda = 3, L = 2, U = 7, shift = 1)
```

* Quantile function of shifted truncated and truncated shifted Poisson distribution.

```{r}
# truncated shifted
qtspois <- qtruncate("spois")
# shifted truncated 
qstpois <- qshift("tpois")
```

```{r}
x <- 1:999/1000
std_distr <- qpois(p = x, lambda = 3)
st_distr <- qstpois(p = x, lambda = 3, L = 2, U = 7, shift = 1)
ts_distr <- qtspois(p = x, lambda = 3, L = 2, U = 7, shift = 1)
```

* random generation function of truncated-shifted normal distribution and of shifted-truncated poisson distribution:

```{r}
# truncated shifted
rtspois <- rtruncate("spois")
# shifted truncated 
rstpois <- rshift("tpois")
```

```{r}
std_rand <- rpois(n=1000, lambda = 3)
st_rand <- rstpois(n=1000, lambda = 3, L = 2, U = 7, shift = 1)
ts_rand <- rtspois(n=1000, lambda = 3, L = 2, U = 7, shift = 1)
```

pl <- ggplot(data = df, mapping = aes(x=value, group=type, fill=type)) +
        geom_density(alpha=0.6) +
        scale_fill_manual(values = c("std_rand" = "cyan3",
                              "st_rand" = "#89FF99",
                              "ts_rand" = "#FF89B4")) +
        ggtitle("Density plot for shifted and not-shifted Normal distributions") +
        theme(legend.title=element_blank(), legend.text= element_text(size=10),
        plot.title = element_text(size=14), axis.title=element_text(size=12))
